%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Shell commands
%
\chapter{Shell commands}
\label{chapter:shell}
\cutname{omake-shell.html}

Shell commands, this is, commands to be executed by the operating system, can be freely mixed with
other code.  The syntax of shell commands is similar but not identical to the syntax used by the
Un*x-shell~\Cmd{bash}{1}.

\textbf{NOTE}: the syntax and shell usage is identical on all platforms, including Win32.  To avoid
portability problems on Win32, it is recommended that you avoid the use of the native shell
interpreter~\Cmd{cmd}{1}.

\begin{verbatim}
    LIB = $(dir lib)
    println(The contents of the $(LIB) directory is:)
    ls $(LIB)
\end{verbatim}

Every command has an integer exit code, which is zero or some other integer.  A command is said to
\emph{succeed} if its exit code is zero.  If a command terminates with a non-zero exit code,
\Cmd{osh}{1} considers the execution to have failed and tells \OMake{} to abort the current process
with an error.

\section{Simple commands}

A simple command is specified with the name of an executable optionally followed by arguments passed
to this executable.  Here are some examples:

\begin{verbatim}
    ls
    ls -AF .
    echo Hello world
\end{verbatim}

The command is found using the current search path in the variable~\verb+PATH[]+, which should
define an array of directories containing the favored executables.

A command may be prefixed by environment variable definitions with the help of the
utility~\Cmd{env}{1}.

\begin{verbatim}
    # Prints "Hello world"
    env X="Hello world" Y=2 printenv X
    # Pass the include path to the Visual C++
    env include="c:\Program Files\Microsoft SDK\include" cl foo.cpp
\end{verbatim}

\section{Globbing}

Commands may contain wildcard patterns.  A pattern specifies a set of files through a limited kind
of regular expression.  Patterns are expanded before the function is executed.

\begin{verbatim}
   # List all files with a .c suffix
   ls *.c

   # List all files with a single character prefix, and .c suffix
   ls ?.c

   # Rename the file hello.ml to foo.ml
   mv {hello,foo}.ml
\end{verbatim}

A comprehensive description of \OMake~glob patterns is given in Section~\ref{section:globbing}.

\section{Background jobs}

A command may also be placed in the background by placing an ampersand~(\verb+&+) after the command.
Control immediately returns to the shell without waiting for the job to complete.  The job continues
to run in the background.

\begin{verbatim}
    gcc -o hugeprogram *.c &
\end{verbatim}

In \Cmd{osh}{1} the ampersand acts command terminator, whereas for \Cmd{bash}{1} it is a command
separator.

\section{Command sequence}\label{section:command-sequence}

Sequence commands by separating them with a semi-colon~(\verb+;+).  The commands get executed from
left to right.  The exit code of the whole sequence is the exit code of the \emph{last} command
executed.  The exit codes of all other commands are \emph{ignored}.

Notes:

\begin{itemize}
  \item In \Cmd{osh}{1} the semicolon strictly acts as a separator.
  \item Each command in a sequence is executed in its own sub-shell.
  \item The property of ignoring all but the last command's exit code can be used to simulate
    GNU~Make's \verb+-+~command-prefix in recipes.
\begin{verbatim}
# GNU Make: ignore exit code of toposort
tsort:
        - toposort --in-place ids.list
\end{verbatim}
    becomes
\begin{verbatim}
# OMake
tsort:
        toposort --in-place ids.list; true
\end{verbatim}
\end{itemize}

\section{File redirection}

The input and the output of a command can be redirected from and to files by adding redirection
operators after the command.

\begin{description}
\item[Redirect input] \Prog{command} \verb+<+ \File{input-file}.

\item[Redirect output] \Prog{command} \verb+>+ \File{output-file}, \Prog{command} \verb+>>+
  \File{output-file}.  The first form truncates \File{output-file}, the
  second form appends to it.

\item[Redirect output and error messages] \Prog{command} \verb+>&+ \File{output-file},
  \Prog{command} \verb+>>&+ \File{output-file}.  Again, the first form truncates \File{output-file}
  and the second form appends to it.
\end{description}

Some examples:

\begin{verbatim}
    # Write to the "foo" file
    echo Hello world > foo

    # Redirect input from the foo file
    cat < foo

    # Redirect standard output and errors to the foo file
    gcc -o boo *.c >& foo
\end{verbatim}

\section{Pipelines}

Pipelines are sequences of commands, where the output of the command on the left-hand side of the
pipe~operator (\verb+|+ and \verb+|&+) is sent to the input of the command on the right-hand side.
With \verb+|+ the output is redirected, but errors are not; with \verb+|&+ both output and errors
are redirected.  Each command in a pipeline is executed in its own sub-shell.

\begin{verbatim}
   # Send the output of the ls command to the printer
   ls *.c | lpr

   # Send output and errors to jyh as email
   gcc -o hugefile *.c |& mail jyh
\end{verbatim}

The pipeline's exit code is the value of the rightmost command to exit with a non-zero code, or zero
if all commands exit successfully.  Note that this behavior is different from ordinary command
sequences (see Section~\ref{section:command-sequence}) and from most other Un*x~shells (those which
supply \verb+pipefail+ set it to \verb+false+ by default), but desirable inside
\OMake~recipes.

\section{Conditional execution}

Commands may also be composed though conditional evaluation using the \verb+||+ and \verb+&&+
syntax.  The expression \verb+command1 && command2+ executes \verb+command2+ only if \verb+command1+
succeeds.  The expression \verb+command1 || command2+ executes \verb+command2+ only if
\verb+command1+ fails.

\begin{verbatim}
   # Display the x/y file if possible
   cd x && cat y

   # Run foo.exe, or print an error message
   (test -x foo.exe && foo.exe) || echo "foo.exe is not executable"
\end{verbatim}

\section{Grouping}

Parenthesis are used for grouping in a pipeline or conditional command.  The grouped commands are
executed in a separate sub-shell.

In the following expression, the \verb+test+~function is used to test whether the \verb+foo.exe+
file is executable.  If it is, the \verb+foo.exe+ file is executed.  If the file is not executable
(or if the \verb+foo.exe+ command fails), the message \verb+"foo.exe is not executable"+ is printed.

\begin{verbatim}
   # Run foo.exe, or print an error message
   (test -x foo.exe && foo.exe) || echo "foo.exe is not executable"
\end{verbatim}

Currently \Cmd{osh}{1} does not support grouping within the current shell as does Bash with curly
braces~(\verb+{}+).

\section{What is a shell command?}

Syntactically, shell commands are any line that is not one of the following:

\begin{itemize}
\item A variable definition of the form \verb+VAR=string+
\item A function call \verb+f(...)+ or method call \verb+o.f(...)+
\item A rule definition containing a colon \verb+string: ...+
\item A special command, including the following:
\begin{itemize}
\item \verb+if ...+
\item \verb+switch ...+
\item \verb+match ...+
\item \verb+section ...+
\item \verb+return ...+
\end{itemize}
\end{itemize}

Commands may also be builtin (aliases).  See the documentation for the
\hyperobj{Shell} for more information.

% -*-
% Local Variables:
% Mode: LaTeX
% fill-column: 100
% TeX-master: "paper"
% TeX-command-default: "LaTeX/dvips Interactive"
% End:
% -*-
