%
%
%
\chapter{Additional build examples}
\label{chapter:build-examples}
\index{build model}
\cutname{omake-build-examples.html}

Let's explain the \OMake{} build model a bit more.
One issue that dominates this discussion is that \OMake{} is based on global project analysis.  That
means you define a configuration for the \emph{entire} project, and you run \emph{one} instance of omake.

For single-directory projects this doesn't mean much.  For multi-directory projects it means a lot.
With GNU make, you would usually invoke the \verb+make+ program recursively for each directory in
the project.  For example, suppose you had a project with some project root directory, containing a
directory of sources \verb+src+, which in turn contains subdirectories \verb+lib+ and \verb+main+.
So your project looks like this nice piece of ASCII art.

\begin{verbatim}
    my_project/
    |--> Makefile
    `--> src/
         |---> Makefile
         |---> lib/
         |     |---> Makefile
         |     `---> source files...
         `---> main/
               |---> Makefile
               `---> source files...
\end{verbatim}
                    
Typically, with GNU make, you would start an instance of \verb+make+ in \verb+my_project/+; this
would in term start an instance of \verb+make+ in the \verb+src/+ directory; and this would start
new instances in \verb+lib/+ and \verb+main/+.  Basically, you count up the number of
\verb+Makefile+s in the project, and that is the number of instances of \verb+make+ processes that
will be created.

The number of processes is no big deal with today's machines (sometimes contrary the the author's opinion, we
no longer live in the 1970s).  The problem with the scheme was that each \verb+make+ process had a
separate configuration, and it took a lot of work to make sure that everything was consistent.
Furthermore, suppose the programmer runs \verb+make+ in the \verb+main/+ directory, but the
\verb+lib/+ is out-of-date.  In this case, \verb+make+ would happily crank away, perhaps trying to
rebuild files in \verb+lib/+, perhaps just giving up.

With \OMake{} this changes entirely.  Well, not entirely.  The source structure is quite similar, we
merely add some Os to the ASCII art.

\begin{verbatim}
    my_project/
    |--> OMakeroot   (or Root.om)
    |--> OMakefile
    `--> src/
         |---> OMakefile
         |---> lib/
         |     |---> OMakefile
         |     `---> source files...
         `---> main/
               |---> OMakefile
               `---> source files...
\end{verbatim}

\index{OMakefile}
\index{OMakeroot}
The role of each \verb+<dir>/OMakefile+ plays the same role as each \verb+<dir>/Makefile+: it
describes how to build the source files in \verb+<dir>+.  The OMakefile retains much of syntax and
structure of the Makefile, but in most cases it is much simpler.

One minor difference is the presence of the OMakeroot in the project root.  The main purpose of this
file is to indicate where the project root \emph{is} in the first place (in case \verb+omake+ is
invoked from a subdirectory).  The \verb+OMakeroot+ serves as the bootstrap file; omake starts by
reading this file first.  Otherwise, the syntax and evaluation of \verb+OMakeroot+ is no different
from any other \verb+OMakefile+.

The \emph{big} difference is that \OMake{} performs a \emph{global} analysis.  Here is what happens
when \verb+omake+ starts.

\index{.SUBDIRS}
\begin{enumerate}
\item omake locates that OMakeroot file, and reads it.
\item Each OMakefile points to its subdirectory OMakefiles using the .SUBDIRS target.
For example, \verb+my_project/OMakefile+ has a rule,

\begin{verbatim}
    .SUBDIRS: src
\end{verbatim}

and the \verb+my_project/src/OMakefile+ has a rule,

\begin{verbatim}
    .SUBDIRS: lib main
\end{verbatim}

\verb+omake+ uses these rules to read and evaluate every \verb+OMakefile+ in the project.
Reading and evaluation is fast.  This part of the process is cheap.

\item Now that the entire configuration is read, \verb+omake+ determines which files are out-of-date
(using a global analysis), and starts the build process.  This may take a while, depending on what
exactly needs to be done.
\end{enumerate}

There are several advantages to this model.  First, since analysis is global, it is much easier to
ensure that the build configuration is consistent--after all, there is only one configuration.
Another benefit is that the build configuration is inherited, and can be re-used, down the
hierarchy.  Typically, the root \verb+OMakefile+ defines some standard boilerplate and
configuration, and this is inherited by subdirectories that tweak and modify it (but do not need to
restate it entirely).  The disadvantage of course is space, since this is global analysis after all.
In practice rarely seems to be a concern; omake takes up much less space than your web browser even
on large projects.

Some notes to the GNU/BSD make user.
\begin{itemize}
\item OMakefiles are a lot like Makefiles.  The syntax is similar, and there many of the builtin
functions are similar.  However, the two build systems are not the same.  Some evil features (in the authors'
opinions) have been dropped in \OMake{}, and some new features have been added.

\item \OMake{} works the same way on all platforms, including Win32.  The standard configuration does
the right thing, but if you care about porting your code to multiple platforms, and you use some
tricky features, you may need to condition parts of your build config on the \verb+$(OSTYPE)+
variable.

\item A minor issue is that \OMake{} dependency analysis is based on MD5 file digests.  That is,
dependencies are based on file \emph{contents}, not file \emph{modification times}.  Say goodbye to
false rebuilds based on spurious timestamp changes and mismatches between local time and fileserver
time.
\end{itemize}

\section{OMakeroot vs. OMakefile}

Before we begin with examples, let's ask the first question, ``What is the difference between the
project root OMakeroot and OMakefile?''  A short answer is, there is no difference, but you must
have an OMakeroot file (or Root.om file).

However, the normal style is that OMakeroot is boilerplate and is more-or-less the same for all
projects.  The OMakefile is where you put all your project-specific stuff.

To get started, you don't have to do this yourself.  In most cases you just perform the following
step in your project root directory.

\begin{itemize}
\item Run \verb+omake --install+ in your project root.
\end{itemize}

This will create the initial OMakeroot and OMakefile files that you can edit to get started.

\section{An example C project}

To begin, let's start with a simple example.  Let's say that we have a full directory tree,
containing the following files.

\begin{verbatim}
    my_project/
    |--> OMakeroot
    |--> OMakefile
    `--> src/
         |---> OMakefile
         |---> lib/
         |     |---> OMakefile
         |     |---> ouch.c
         |     |---> ouch.h
         |     `---> bandaid.c
         `---> main/
               |---> OMakefile
               |---> horsefly.c
               |---> horsefly.h
               `---> main.c
\end{verbatim}

Here is an example listing.

\begin{verbatim}
my_project/OMakeroot:
    # Include the standard configuration for C applications
    open build/C
    
    # Process the command-line vars
    DefineCommandVars()
    
    # Include the OMakefile in this directory.
    .SUBDIRS: .

my_project/OMakefile:
    # Set up the standard configuration
    CFLAGS += -g

    # Include the src subdirectory
    .SUBDIRS: src

my_project/src/OMakefile:
    # Add any extra options you like
    CFLAGS += -O2

    # Include the subdirectories
    .SUBDIRS: lib main

my_project/src/lib/OMakefile:
    # Build the library as a static library.
    # This builds libbug.a on Unix/OSX, or libbug.lib on Win32.
    # Note that the source files are listed _without_ suffix.
    StaticCLibrary(libbug, ouch bandaid)

my_project/src/main/OMakefile:
    # Some files include the .h files in ../lib
    INCLUDES += ../lib

    # Indicate which libraries we want to link against.
    LIBS[] +=
        ../lib/libbug

    # Build the program.
    # Builds horsefly.exe on Win32, and horsefly on Unix.
    # The first argument is the name of the executable.
    # The second argument is an array of object files (without suffix)
    # that are part of the program.
    CProgram(horsefly, horsefly main)

    # Build the program by default (in case omake is called
    # without any arguments).  EXE is defined as .exe on Win32,
    # otherwise it is empty.
    .DEFAULT: horsefly$(EXE)
\end{verbatim}

Most of the configuration here is defined in the file \verb+build/C.om+ (which is part of the \OMake{}
distribution).  This file takes care of a lot of work, including:
\begin{itemize}
\item Defining the \verb+StaticCLibrary+ and \verb+CProgram+ functions, which describe the canonical
way to build C libraries and programs.
\item Defining a mechanism for \emph{scanning} each of the source programs to discover dependencies.
That is, it defines .SCANNER rules for C source files.
\end{itemize}

Variables are inherited down the hierarchy, so for example, the value of CFLAGS in
src/main/OMakefile is ``\verb+-g -O2+''.

\section{An example OCaml project}

Let's repeat the example, assuming we are using OCaml instead of C.
This time, the directory tree looks like this.

\begin{verbatim}
    my_project/
    |--> OMakeroot
    |--> OMakefile
    `--> src/
         |---> OMakefile
         |---> lib/
         |     |---> OMakefile
         |     |---> ouch.ml
         |     |---> ouch.mli
         |     `---> bandaid.ml
         `---> main/
               |---> OMakefile
               |---> horsefly.ml
               |---> horsefly.mli
               `---> main.ml
\end{verbatim}

The listing is only a bit different.

\begin{verbatim}
my_project/OMakeroot:
    # Include the standard configuration for OCaml applications
    open build/OCaml

    # Process the command-line vars
    DefineCommandVars()

    # Include the OMakefile in this directory.
    .SUBDIRS: .

my_project/OMakefile:
    # Set up the standard configuration
    OCAMLFLAGS += -Wa

    # Do we want to use the bytecode compiler,
    # or the native-code one?  Let's use both for
    # this example.
    NATIVE_ENABLED = true
    BYTE_ENABLED = true

    # Include the src subdirectory
    .SUBDIRS: src

my_project/src/OMakefile:
    # Include the subdirectories
    .SUBDIRS: lib main

my_project/src/lib/OMakefile:
    # Let's do aggressive inlining on native code
    OCAMLOPTFLAGS += -inline 10

    # Build the library as a static library.
    # This builds libbug.a on Unix/OSX, or libbug.lib on Win32.
    # Note that the source files are listed _without_ suffix.
    OCamlLibrary(libbug, ouch bandaid)

my_project/src/main/OMakefile:
    # These files depend on the interfaces in ../lib
    OCAMLINCLUDES += ../lib

    # Indicate which libraries we want to link against.
    OCAML_LIBS[] +=
        ../lib/libbug

    # Build the program.
    # Builds horsefly.exe on Win32, and horsefly on Unix.
    # The first argument is the name of the executable.
    # The second argument is an array of object files (without suffix)
    # that are part of the program.
    OCamlProgram(horsefly, horsefly main)

    # Build the program by default (in case omake is called
    # without any arguments).  EXE is defined as .exe on Win32,
    # otherwise it is empty.
    .DEFAULT: horsefly$(EXE)
\end{verbatim}

In this case, most of the configuration here is defined in the file \verb+build/OCaml.om+.  In this
particular configuration, files in \verb+my_project/src/lib+ are compiled aggressively with the
option \verb+-inline 10+, but files in \verb+my_project/src/lib+ are compiled normally.

\section{OCaml~Library with C~Dependencies}

Build the OCaml~library~\verb+libstring_extra+, which depends on both,
OCaml~code and C~code.

\begin{verbatim}
    lib/
    |---> OMakefile
    |---> string_extra.mli
    |---> string_extra.ml
    |---> index_from_opt_interface.c
\end{verbatim}

This is the OCaml interface file~\verb+string_extra.mli+:

\begin{verbatim}
(** [index_from_opt a_haystack an_initial_index a_needle]
 *  @raise Invalid_argument if the initial index is out of range. *)
val index_from_opt: string -> int -> string -> int option

(** [index_from a_haystack an_initial_index a_needle]
 *  @raise Invalid_argument if the initial index is out of range.
 *  @raise Not_found if there is no needle in haystack after the initial index. *)
val index_from: string -> int -> string -> int
\end{verbatim}

The associated OCaml code file~\verb+string_extra.ml+ looks like this:

\begin{verbatim}
external index_from_opt: string -> int -> string -> int option = "index_from_opt_interface"

let index_from a_haystack an_initial_index a_needle =
  match index_from_opt a_haystack an_initial_index a_needle with
    None -> raise Not_found
  | Some index -> index
\end{verbatim}

The C~code in \verb+index_from_opt_interface.c+ is gross because we
must play nice with OCaml's garbage collector.  We use
function~\verb+memmem+ of the GNU~C-library because OCaml~strings are
not null-terminated.

\begin{verbatim}
#define _GNU_SOURCE           /* Pull memmem(3) into scope. */
#include <string.h>           /* NULL, size_t; _GNU_SOURCE: memmem() */

#define CAML_NAME_SPACE       /* Only define objects with "caml_" prefix. */
#include <caml/alloc.h>       /* caml_alloc() */
#include <caml/fail.h>        /* caml_invalid_argument() */
#include <caml/memory.h>      /* CAMLlocal1(), CAMLparam1(), CAMLparam3(),
                               * CAMLreturn(), Store_field() */
#include <caml/mlvalues.h>    /* Long_val(), String_val(), Val_int(),
                               * caml_string_length(), value */

#define Val_none Val_int(0)


inline static
value
Val_some(value a_value)
{
    CAMLparam1(a_value);
    CAMLlocal1(some);

    some = caml_alloc(1, 0);
    Store_field(some, 0, a_value);

    CAMLreturn(some);
}


CAMLprim
value
index_from_opt_interface(value a_haystack, value an_initial_index, value a_needle)
{
    CAMLparam3(a_haystack, an_initial_index, a_needle);

    const char *const haystack = String_val(a_haystack);
    const size_t haystack_length = caml_string_length(a_haystack);
    const long initial_index = Long_val(an_initial_index);

    if (initial_index >= 0L && (size_t) initial_index < haystack_length)
    {
        const char *const index =
             memmem(haystack + initial_index,
                    caml_string_length(a_haystack) - (size_t) initial_index,
                    String_val(a_needle),
                    caml_string_length(a_needle));

        CAMLreturn((index == NULL) ? Val_none : Val_some(Val_int(index - haystack)));
    }
    else
    {
        caml_invalid_argument("index_from_opt_interface");
    }
}
\end{verbatim}

The \verb+OMakefile+ however is very simple; we just have to add the
path to the OCaml~compiler's interface files to \verb+CFLAGS+.

\begin{verbatim}
OCAMLFLAGS = -g -strict-formats -strict-sequence -w +a
CFLAGS = -O2 -g -Wall -Wextra -Wconversion -I$(OCAMLLIB)

OCamlMixedLibrary(libstring_extra, string_extra, index_from_opt_interface)
\end{verbatim}

To use the library put \verb+OCAML_LIBS = libstring_extra+ in the
relevant \verb+OMakefile+.

\section{Handling new languages}
\index{cats and dogs}

The previous two examples seem to be easy enough, but they rely on the \OMake{} standard library (the
files \verb+build/C+ and \verb+build/OCaml+) to do all the work.  What happens if we want to write a
build configuration for a language that is not already supported in the \OMake{} standard library?

For this example, let's suppose we are adopting a new language.  The language uses the standard
compile/link model, but is not in the \OMake{} standard library.  Specifically, let's say we have the
following setup.

\begin{itemize}
\item Source files are defined in files with a \verb+.cat+ suffix (for Categorical Abstract Terminology).
\item \verb+.cat+ files are compiled with the \verb+catc+ compiler to produce \verb+.woof+ files
(Wicked Object-Oriented Format).
\item \verb+.woof+ files are linked by the \verb+catc+ compiler with the \verb+-c+ option to produce
a \verb+.dog+ executable (Digital Object Group).  The \verb+catc+ also defines a \verb+-a+ option to
combine several \verb+.woof+ files into a library.
\item Each \verb+.cat+ can refer to other source files.  If a source file \verb+a.cat+ contains a
line \verb+open b+, then \verb+a.cat+ depends on the file \verb+b.woof+, and \verb+a.cat+ must be
recompiled if \verb+b.woof+ changes.  The \verb+catc+ function takes a \verb+-I+ option to define a
search path for dependencies.
\end{itemize}

To define a build configuration, we have to do three things.
\begin{enumerate}
\item Define a \verb+.SCANNER+ rule for discovering dependency information for the source files.
\item Define a generic rule for compiling a \verb+.cat+ file to a \verb+.woof+ file.
\item Define a rule (as a function) for linking \verb+.woof+ files to produce a \verb+.dog+ executable.
\end{enumerate}

Initially, these definitions will be placed in the project root \verb+OMakefile+.

\subsection{Defining a default compilation rule}

Let's start with part 2, defining a generic compilation rule.  We'll define the build rule as an
\emph{implicit} rule.  To handle the include path, we'll define a variable \verb+CAT_INCLUDES+ that
specifies the include path.  This will be an array of directories.  To define the options, we'll use
a lazy variable (Section~\ref{section:lazy}).  In case there
are any other standard flags, we'll define a \verb+CAT_FLAGS+ variable.

\begin{verbatim}
   # Define the catc command, in case we ever want to override it
   CATC = catc

   # The default flags are empty
   CAT_FLAGS =
   
   # The directories in the include path (empty by default)
   INCLUDES[] =

   # Compute the include options from the include path
   PREFIXED_INCLUDES[] = $`(mapprefix -I, $(INCLUDES))

   # The default way to build a .woof file
   %.woof: %.cat
       $(CATC) $(PREFIXED_INCLUDES) $(CAT_FLAGS) -c $<
\end{verbatim}

The final part is the build rule itself, where we call the \verb+catc+ compiler with the include
path, and the \verb+CAT_FLAGS+ that have been defined.  The \verb+$<+ variable represents the source
file.

\subsection{Defining a rule for linking}

For linking, we'll define another rule describing how to perform linking.  Instead of defining an
implicit rule, we'll define a function that describes the linking step.  The function will take two
arguments; the first is the name of the executable (without suffix), and the second is the files to
link (also without suffixes).  Here is the code fragment.

\begin{verbatim}
    # Optional link options
    CAT_LINK_FLAGS =

    # The function that defines how to build a .dog program
    CatProgram(program, files) =
        # Add the suffixes
        file_names = $(addsuffix .woof, $(files))
        prog_name = $(addsuffix .dog, $(program))

        # The build rule
        $(prog_name): $(file_names)
            $(CATC) $(PREFIXED_INCLUDES) $(CAT_FLAGS) $(CAT_LINK_FLAGS) -o $@ $+

        # Return the program name
        value $(prog_name)
\end{verbatim}

The \verb+CAT_LINK_FLAGS+ variable is defined just in case we want to pass additional flags specific
to the link step.  Now that this function is defined, whenever we want to define a rule for building
a program, we simply call the rule.  The previous implicit rule specifies how to compile each source file,
and the \verb+CatProgram+ function specifies how to build the executable.

\begin{verbatim}
    # Build a rover.dog program from the source
    # files neko.cat and chat.cat.
    # Compile it by default.
    .DEFAULT: $(CatProgram rover, neko chat)
\end{verbatim}

\subsection{Dependency scanning}
\label{section:scanner-exm}\index{.SCANNER}

That's it, almost.  The part we left out was automated dependency scanning.  This is one of the
nicer features of \OMake{}, and one that makes build specifications easier to write and more robust.
Strictly speaking, it isn't required, but you definitely want to do it.

The mechanism is to define a \verb+.SCANNER+ rule, which is like a normal rule, but it specifies how
to compute dependencies, not the target itself.  In this case, we want to define a \verb+.SCANNER+
rule of the following form.

\begin{verbatim}
    .SCANNER: %.woof: %.cat
        <commands>
\end{verbatim}

This rule specifies that a \verb+.woof+ file may have additional dependencies that can be extracted
from the corresponding \verb+.cat+ file by executing the \verb+<commands>+.  The \emph{result} of
executing the \verb+<commands>+ should be a sequence of dependencies in \OMake{} format, printed to the
standard output.

As we mentioned, each \verb+.cat+ file specifies dependencies on \verb+.woof+ files with an
\verb+open+ directive.  For example, if the \verb+neko.cat+ file contains a line \verb+open chat+,
then \verb+neko.woof+ depends on \verb+chat.woof+.  In this case, the \verb+<commands>+ should print
the following line.

\begin{verbatim}
    neko.woof: chat.woof
\end{verbatim}

For an analogy that might make this clearer, consider the C programming language, where a \verb+.o+
file is produced by compiling a \verb+.c+ file.  If a file \verb+foo.c+ contains a line like
\verb+#include "fum.h"+, then \verb+foo.c+ should be recompiled whenever \verb+fum.h+ changes.  That
is, the file \verb+foo.o+ \emph{depends} on the file \verb+fum.h+.  In the \OMake{} parlance, this is
called an \emph{implicit} dependency, and the \verb+.SCANNER+ \verb+<commands>+ would print a line
like the following.

\begin{verbatim}
    foo.o: fum.h
\end{verbatim}

\index{awk} Now, returning to the animal world, to compute the dependencies of \verb+neko.woof+, we
should scan \verb+neko.cat+, line-by-line, looking for lines of the form \verb+open <name>+.  We
could do this by writing a program, but it is easy enough to do it in \verb+omake+ itself.  We can
use the builtin \hyperfun{awk} to scan the source file.  One slight complication
is that the dependencies depend on the \verb+INCLUDE+ path.  We'll use the
\hyperfun{find-in-path} to find them.  Here we go.

\begin{verbatim}
    .SCANNER: %.woof: %.cat
        section
            # Scan the file
            deps[] =
            awk($<)
            case $'^open'
                deps[] += $2
                export

            # Remove duplicates, and find the files in the include path
            deps = $(find-in-path $(INCLUDES), $(set $(deps)))

            # Print the dependencies
            println($"$@: $(deps)")
\end{verbatim}

Let's look at the parts.  First, the entire body is defined in a \verb+section+ because we are
computing it internally, not as a sequence of shell commands.

We use the \verb+deps+ variable to collect all the dependencies.  The \verb+awk+ function scans the
source file (\verb+$<+) line-by-line.  For lines that match the regular expression \verb+^open+
(meaning that the line begins with the word \verb+open+), we add the second word on the line to the
\verb+deps+ variable.  For example, if the input line is \verb+open chat+, then we would add the
\verb+chat+ string to the \verb+deps+ array.  All other lines in the source file are ignored.

Next, the \verb+$(set $(deps))+ expression removes any duplicate values in the \verb+deps+ array
(sorting the array alphabetically in the process).  The \verb+find-in-path+ function then finds the
actual location of each file in the include path.

The final step is print the result as the string \verb+$"$@: $(deps)"+ The quotations are added to
flatten the \verb+deps+ array to a simple string.

\subsection{Pulling it all together}

To complete the example, let's pull it all together into a single project, much like our previous
example.

\begin{verbatim}
    my_project/
    |--> OMakeroot
    |--> OMakefile
    `--> src/
         |---> OMakefile
         |---> lib/
         |     |---> OMakefile
         |     |---> neko.cat
         |     `---> chat.cat
         `---> main/
               |---> OMakefile
               `---> main.cat
\end{verbatim}

The listing for the entire project is as follows.  Here, we also include a function
\verb+CatLibrary+ to link several \verb+.woof+ files into a library.

\begin{verbatim}
my_project/OMakeroot:
    # Process the command-line vars
    DefineCommandVars()
    
    # Include the OMakefile in this directory.
    .SUBDIRS: .

my_project/OMakefile:
   ########################################################################
   # Standard config for compiling .cat files
   #

   # Define the catc command, in case we ever want to override it
   CATC = catc

   # The default flags are empty
   CAT_FLAGS =
   
   # The directories in the include path (empty by default)
   INCLUDES[] =

   # Compute the include options from the include path
   PREFIXED_INCLUDES[] = $`(mapprefix -I, $(INCLUDES))

   # Dependency scanner for .cat files
   .SCANNER: %.woof: %.cat
        section
            # Scan the file
            deps[] =
            awk($<)
            case $'^open'
                deps[] += $2
                export

            # Remove duplicates, and find the files in the include path
            deps = $(find-in-path $(INCLUDES), $(set $(deps)))

            # Print the dependencies
            println($"$@: $(deps)")

   # The default way to compile a .cat file
   %.woof: %.cat
       $(CATC) $(PREFIXED_INCLUDES) $(CAT_FLAGS) -c $<

   # Optional link options
   CAT_LINK_FLAGS =

   # Build a library for several .woof files
   CatLibrary(lib, files) =
       # Add the suffixes
       file_names = $(addsuffix .woof, $(files))
       lib_name = $(addsuffix .woof, $(lib))

       # The build rule
       $(lib_name): $(file_names)
           $(CATC) $(PREFIXED_INCLUDES) $(CAT_FLAGS) $(CAT_LINK_FLAGS) -a $@ $+

       # Return the program name
       value $(lib_name)

   # The function that defines how to build a .dog program
   CatProgram(program, files) =
       # Add the suffixes
       file_names = $(addsuffix .woof, $(files))
       prog_name = $(addsuffix .dog, $(program))

       # The build rule
       $(prog_name): $(file_names)
           $(CATC) $(PREFIXED_INCLUDES) $(CAT_FLAGS) $(CAT_LINK_FLAGS) -o $@ $+

       # Return the program name
       value $(prog_name)

   ########################################################################
   # Now the program proper
   #

   # Include the src subdirectory
   .SUBDIRS: src

my_project/src/OMakefile:
   .SUBDIRS: lib main

my_project/src/lib/OMakefile:
   CatLibrary(cats, neko chat)

my_project/src/main/OMakefile:
   # Allow includes from the ../lib directory
   INCLUDES[] += ../lib

   # Build the program
   .DEFAULT: $(CatProgram main, main ../cats)
\end{verbatim}

Some notes.  The configuration in the project \verb+OMakeroot+ defines the standard configuration, including
the dependency scanner, the default rule for compiling source files, and functions for building
libraries and programs.

These rules and functions are inherited by subdirectories, so the \verb+.SCANNER+ and build rules
are used automatically in each subdirectory, so you don't need to repeat them.

\subsection{Finishing up}

At this point we are done, but there are a few things we can consider.

First, the rules for building cat programs is defined in the project \verb+OMakefile+.  If you had
another cat project somewhere, you would need to copy the \verb+OMakeroot+ (and modify it as
needed).  Instead of that, you should consider moving the configuration to a shared library
directory, in a file like \verb+Cat.om+.  That way, instead of copying the code, you could include
the shared copy with an \OMake{} command \verb+open Cat+.  The share directory should be added to your
\verb+OMAKEPATH+ environment variable to ensure that \verb+omake+ knows how to find it.

Better yet, if you are happy with your work, consider submitting it as a standard configuration (by
sending a request to \verb+omake@metaprl.org+) so that others can make use of it too.

\section{Collapsing the hierarchy, .SUBDIRS bodies}
\index{.SUBDIRS bodies}

Some projects have many subdirectories that all have the same configuration.  For instance, suppose
you have a project with many subdirectories, each containing a set of images that are to be composed
into a web page.  Apart from the specific images, the configuration of each file is the same.

To make this more concrete, suppose the project has four subdirectories \verb+page1+, \verb+page2+,
\verb+page3+, and \verb+page4+.  Each contains two files \verb+image1.jpg+ and \verb+image2.jpg+
that are part of a web page generated by a program \verb+genhtml+.

Instead of of defining a \verb+OMakefile+ in each directory, we can define it as a body to the
\verb+.SUBDIRS+ command.

\begin{verbatim}
    .SUBDIRS: page1 page2 page3 page4
        index.html: image1.jpg image2jpg
            genhtml $+ > $@
\end{verbatim}

The body of the \verb+.SUBDIRS+ is interpreted exactly as if it were the \verb+OMakefile+, and it
can contain any of the normal statements.  The body is evaluated \emph{in the subdirectory} for each
of the subdirectories.  We can see this if we add a statement that prints the current directory
(\verb+$(CWD)+).

\begin{verbatim}
    .SUBDIRS: page1 page2 page3 page4
        println($(absname $(CWD)))
        index.html: image1.jpg image2jpg
            genhtml $+ > $@
  # prints
    /home/jyh/.../page1
    /home/jyh/.../page2
    /home/jyh/.../page3
    /home/jyh/.../page4
\end{verbatim}

\subsection{Using glob patterns}

Of course, this specification is quite rigid.  In practice, it is likely that each subdirectory will
have a different set of images, and all should be included in the web page.  One of the easier
solutions is to use one of the directory-listing functions, like
\hyperfunn{glob} or \hyperfunn{ls}.
The \verb+glob+ function takes a shell pattern, and returns an array of
file with matching filenames in the current directory.

\begin{verbatim}
    .SUBDIRS: page1 page2 page3 page4
        IMAGES = $(glob *.jpg)
        index.html: $(IMAGES)
            genhtml $+ > $@
\end{verbatim}

\subsection{Simplified sub-configurations}

Another option is to add a configuration file in each of the subdirectories that defines
directory-specific information.  For this example, we might define a file \verb+BuildInfo.om+ in
each of the subdirectories that defines a list of images in that directory.  The \verb+.SUBDIRS+
line is similar, but we include the BuildInfo file.

\begin{verbatim}
    .SUBDIRS: page1 page2 page3 page4
        include BuildInfo   # Defines the IMAGES variable

        index.html: $(IMAGES)
            genhtml $+ > $@
\end{verbatim}

Where we might have the following configurations.

\begin{verbatim}
   page1/BuildInfo.om:
       IMAGES[] = image.jpg
   page2/BuildInfo.om:
       IMAGES[] = ../common/header.jpg winlogo.jpg
   page3/BuildInfo.om:
       IMAGES[] = ../common/header.jpg unixlogo.jpg daemon.jpg
   page4/BuildInfo.om:
       IMAGES[] = fee.jpg fi.jpg foo.jpg fum.jpg
\end{verbatim}

\subsection{Computing the subdirectory list}

The other hardcoded specification is the list of subdirectories \verb+page1+, ..., \verb+page4+.
Rather than editing the project \verb+OMakefile+ each time a directory is added, we could compute it
(again with \verb+glob+).

\begin{verbatim}
    .SUBDIRS: $(glob page*)
        index.html: $(glob *.jpg)
            genhtml $+ > $@
\end{verbatim}

Alternately, the directory structure may be hierarchical.  Instead of using \verb+glob+, we could
use the \verb+subdirs+ function, returns each of the directories in a hierarchy.  For example, this
is the result of evaluating the \verb+subdirs+ function in the omake project root.  The \verb+P+
option, passed as the first argument, specifies that the listing is ``proper,'' it should not
include the \verb+omake+ directory itself.

\begin{verbatim}
    osh> subdirs(P, .)
    - : <array
            /home/jyh/.../omake/mk : Dir
            /home/jyh/.../omake/RPM : Dir
            ...
            /home/jyh/.../omake/osx_resources : Dir>
\end{verbatim}

Using \verb+subdirs+, our example is now as follows.

\begin{verbatim}
    .SUBDIRS: $(subdirs P, .)
        index.html: $(glob *.jpg)
            genhtml $+ > $@
\end{verbatim}

In this case, \emph{every} subdirectory will be included in the project.

If we are using the \verb+BuildInfo.om+ option.  Instead of including every subdirectory, we could
include only those that contain a \verb+BuildInfo.om+ file.  For this purpose, we can use the
\verb+find+ function, which traverses the directory hierarchy looking for files that match a test
expression.  In our case, we want to search for files with the name \verb+BuildInfo.om+.
Here is an example call.

\begin{verbatim}
    osh> FILES = $(find . -name BuildInfo.om)
    - : <array
            /home/jyh/.../omake/doc/html/BuildInfo.om : File
            /home/jyh/.../omake/src/BuildInfo.om : File
            /home/jyh/.../omake/tests/simple/BuildInfo.om : File>
    osh> DIRS = $(dirof $(FILES))
    - : <array
            /home/jyh/.../omake/doc/html : Dir
            /home/jyh/.../omake/src : Dir
            /home/jyh/.../omake/tests/simple : Dir>
\end{verbatim}

In this example, there are three \verb+BuildInfo.om+ files, in the \verb+doc/html+, \verb+src+, and
\verb+tests/simple+ directories.  The \verb+dirof+ function returns the directories for each of the
files.

Returning to our original example, we modify it as follows.

\begin{verbatim}
    .SUBDIRS: $(dirof $(find . -name BuildInfo.om))
        include BuildInfo   # Defines the IMAGES variable

        index.html: $(IMAGES)
            genhtml $+ > $@
\end{verbatim}

\subsection{Temporary directories}

Sometimes, your project may include temporary directories--directories where you place intermediate
results.  these directories are deleted whenever the project is cleanup up.  This means, in
particular, that you can't place an \verb+OMakefile+ in a temporary directory, because it will be
removed when the directory is removed.

Instead, if you need to define a configuration for any of these directories, you will need to define
it using a \verb+.SUBDIRS+ body.

\begin{verbatim}
    section
        CREATE_SUBDIRS = true

        .SUBDIRS: tmp
            # Compute an MD5 digest
            %.digest: %.comments
               echo $(digest $<) > $@

            # Extract comments from the source files
            %.comments: ../src/%.src
               grep '^#' $< > $@

            .DEFAULT: foo.digest

    .PHONY: clean

    clean:
        rm -rf tmp        
\end{verbatim}

In this example, we define the \verb+CREATE_SUBDIRS+ variable as true, so that the \verb+tmp+
directory will be created if it does not exist.  The \verb+.SUBDIRS+ body in this example is a bit
contrived, but it illustrates the kind of specification you might expect.  The \verb+clean+
phony-target indicates that the \verb+tmp+ directory should be removed when the project is cleaned
up.

% -*-
% Local Variables:
% Mode: LaTeX
% fill-column: 100
% TeX-master: "paper"
% TeX-command-default: "LaTeX/dvips Interactive"
% End:
% -*-
